An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-09-28 paper.tex
Update §3.1

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/rfcs_areas.pdf}
  \caption{
    RFCs by area. \textit{Other} includes legacy RFCs, and RFCs from other
    non-IETF streams.
  }
  \label{fig:rfcs_by_area}
\end{figure}

@@

%..................................................................................................
\pb{Role of Working Groups:}
From the creation of the IETF in 1986, the growing community with its
interests in an increasing set of applications and protocols, has been
split into working groups. These working groups are chartered with a focus
on a well-defined programme of work, and exist within areas that have a
broader focus.  As shown in Figure~\ref{fig:rfcs_by_area}, the output of
different areas has remained relatively stable over time. The most notable
trends begin with the creation of the Real-time Applications and
Infrastructure (rai) area from within the Transport (tsv) area, and its
later merger with the Applications (app) area to become the Applications
and Real-Time (art) area around 2014. Additionally, we also observe the
significant growth in output of the Routing (rtg) area, owing to the
ongoing development of standards for MPLS, service function chaining, and
fat tree routing in data centres.

To give a sense for the broader productivity of the IETF,
Figure~\ref{fig:pub_wgs_yearly} shows the number of working groups that
publish RFCs each year. This highlights how the structure of the IETF has
grown to accommodate its larger community: in the early 1990s, fewer than
20 working groups were actively publishing RFCs, while in recent years
there has typically been at least 60 different publishing groups, with a
peak of 97 active working groups, and other activities, in 2011.

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/unique_wgs_per_year_areas.pdf}
  \caption{
    Number of publishing working groups. \textit{Other} includes legacy
    RFCs, IRTF research groups, and non-IETF streams.
  }
  \label{fig:pub_wgs_yearly}
\end{figure}

Figure~\ref{fig:rfcs_days_to_pub} plots the median number of days from the
submission of an RFC's first draft, through to its publication as an RFC.
This shows a clear trend: RFCs are taking longer to make their way through
the standardisation and publication process. The median number of days to
publication was 469 in 2001, rising to 1,170 in 2020. Further,
Figure~\ref{fig:drafts_year} shows the median number of Internet-Drafts
that are posted before an RFC is published. Days to publication and number
of drafts are strongly correlated, suggesting that the time is spent making
changes to the document. This may go some way towards explaining the
decline in output of the IETF: each RFC is taking longer to produce, with
more revisions before publication.

@@

@@

%..................................................................................................
\pb{RFC Length:}
One may conjecture that this slowdown is driven by longer RFCs that contain
more material. To explore this, Figure~\ref{fig:pages_year} shows the
median page count of RFCs. This shows that the increase in the duration of
the standardisation process for RFCs cannot be attributed to RFCs becoming
longer: median page counts have remained stable.

@@

%..................................................................................................
\pb{Relationships between RFCs:}
RFCs themselves may be becoming more complex as a result of them needing to
describe their relationship with older standards. As the Internet has
evolved and matured, applications and protocols must be maintained, and new
standards need to interoperate with existing RFCs.
Figure~\ref{fig:updates_year} highlights this, showing the proportion of
RFCs that are published each year that update (i.e., extend or augment) or
obsolete (i.e., replace) one or more previously published RFCs. Intuitively,
this percentage has slowly increased as the IETF has matured: in 2020, more
than 30\% of RFCs updated or made obsolete a previous RFC.
Figure~\ref{fig:citations_year} expands on this, showing the median number
of citations from each RFC to other Internet-Drafts and RFCs. This
similarly shows that RFCs are increasingly referring to prior work.

@@

%..................................................................................................
\pb{Use of requirements-setting language:}
Figure~\ref{fig:keyword_usage_rates} further confirms the growing
complexity of RFCs, showing how the use of keywords has evolved over time.
Keywords are used in RFCs to indicate the normative requirements an RFC
imposes on implementations. Figure~\ref{fig:keyword_usage_rates} shows the
total number of occurrences of each of the RFC 2119~\cite{RFC2119} keywords
(i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD NOT,
RECOMMENDED, MAY, OPTIONAL), divided by the page count of the RFC. As
shown, the median number of keywords per page grew from 2001 through to
2010, indicating a growing number of requirements being expressed in RFCs,
before plateauing in recent years.

@@

\todo{Should we include ``Academic impact of RFCs'' from IMC 2021?}

%..................................................................................................
\pb{Summary:}
RFCs are taking longer to produce, and they go through a greater number of
revisions before publication. Further, they increasingly update or
reference previously published RFCs, and make greater use of
requirements-setting keywords. Yet, while these measures indicate that the
overall complexity of standards documents is increasing, none of these
factors strongly correlate with the time taken to publish RFCs. This
indicates that these factors aren't driving the increasing duration of the
standardisation process. In the sections that follow, we explore trends in
authorship and community interaction that may also impact the process.

>>>>

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/rfcs_areas.pdf}
  \caption{
    Number of RFCs published over time, subdivided by IETF Area.
    The areas are 
    \emph{art} -- applications and real-time;
    \emph{app} -- applications;
    \emph{int} -- internetworking;
    \emph{ops} -- operations and management;
    \emph{rai} -- real-time applications and infrastructure;
    \emph{rtg} -- routing;
    \emph{sec} -- security; and
    \emph{tsv} -- transport and services.
    The \emph{other} category includes legacy RFCs, pre-dating the creation
    of the IETF, and RFCs published in non-IETF publication streams such as
    the IRTF, IAB, and the Independent stream.
  }
  \label{fig:rfcs_by_area}
\end{figure}

@@

%..................................................................................................
\pb{Role of Working Groups:}
From the creation of the IETF in 1986 the growing community, with its
interests in an increasing set of applications and protocols, has been
split into a number of \emph{working groups}.\footnote{There are 126
active working groups at the time of writing, organised into 7 areas.
The number of groups varies over time, but is typically in the range
100-150.} These working groups are each chartered with a well-defined
programme of work and exist within \emph{areas} that have a broader focus.

As shown in Figure~\ref{fig:rfcs_by_area}, the output of
different areas has remained relatively stable over time. The most notable
trends begin with the creation of the Real-time Applications and
Infrastructure (rai) area from within the Transport (tsv) area, and its
later merger with the Applications (app) area to become the Applications
and Real-Time (art) area around 2014. Additionally, we also observe the
significant growth in output of the Routing (rtg) area, owing to the
ongoing development of standards for MPLS, service function chaining, and
fat tree routing in data centres.

To give a sense for the broader productivity of the IETF,
Figure~\ref{fig:pub_wgs_yearly} shows the number of working groups that
publish RFCs each year. This highlights how the structure of the IETF has
grown to accommodate its larger community: in the early 1990s, fewer than
20 working groups were published an RFC each year, while in recent years
there has typically been at least 60 working group publishing RFCs, with
a peak of 97 working groups publishing RFCs in 2011 (the number of groups
publishing RFCs is less than the total number of groups, because it takes
time for the work of a new group to reach maturity).

\begin{figure}
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/imc-2021/documents/unique_wgs_per_year_areas.pdf}
  \caption{
    Number of groups publishing RFCs over time, subdivided by IETF Area.
    The areas are as described in Figure~\ref{fig:rfcs_by_area}.
  }
  \label{fig:pub_wgs_yearly}
\end{figure}

%..................................................................................................
\pb{Factors Affecting Publication Rate:} 
The standards development process is taking longer.
Figure~\ref{fig:rfcs_days_to_pub} plots the median number of days from the
submission of the first draft of a document through to its publication as
an RFC. This shows a clear trend: RFCs are taking longer to make their way
through the standardisation and publication process. The median number of
days to publication was 469 in 2001, rising to 1,170 in 2020.

@@

Similarly, Figure~\ref{fig:drafts_year} shows the median number of
revisions each document undergoes before being published as an RFC. Days to
publication and number of revisions are strongly correlated, suggesting
that the time is spent making changes to the document. This may go some way
towards explaining the decline in output of the IETF: each RFC is taking
longer to produce, with more revisions before publication.

@@

What factors affect the publication rate?  One may conjecture that this
slowdown is driven by longer RFCs that contain more material. Yet from the
median page count of RFCs, shown in Figure~\ref{fig:pages_year}, no clear
trend being visible. The increase in the duration of the standardisation
process for RFCs cannot be attributed to RFCs becoming longer: page counts
have remained stable.

@@

The publication rate might also be slowing because RFCs are becoming more
complex. This could occur, for example, as a result of having to maintain
compatibility with older RFCs as the Internet has evolved and matured.
Figure~\ref{fig:updates_year} shows that this is the case, plotting the
proportion of RFCs that are published each year that update (i.e., extend
or augment) or obsolete (i.e., replace) one or more previously published
RFCs.  This percentage has slowly increased as the IETF has matured: in
2020, more than 30\% of RFCs updated or made obsolete a previous RFC.
%
Figure~\ref{fig:citations_year} expands on this, showing the median number
of citations from each RFC to other Internet-Drafts and RFCs. This
similarly shows that RFCs are increasingly referring to prior work.

@@

Figure~\ref{fig:keyword_usage_rates} further confirms the growing complexity
of RFCs, showing how the use of normative language keywords has evolved
over time.  Keywords are used in RFCs to indicate the normative requirements
an RFC imposes on implementations. Figure~\ref{fig:keyword_usage_rates}
shows the total number of occurrences of each of the RFC 2119~\cite{RFC2119}
keywords (i.e., MUST, MUST NOT, REQUIRED, SHALL, SHALL NOT, SHOULD, SHOULD
NOT, RECOMMENDED, MAY, OPTIONAL), divided by the page count of the RFC. As
shown, the median number of keywords per page grew from 2001 through to
2010, indicating a growing number of requirements being expressed in RFCs,
before plateauing in recent years.

@@

%..................................................................................................
\pb{Summary:}
RFCs are taking longer to produce, and they go through a greater number
of revisions before publication. Further, they increasingly update or
reference previously published RFCs, and make greater use of
requirements-setting keywords. These are all hallmarks of a maturing
standards development organisation that must increasingly consider
compatibility with the installed based in the development of its standards.
